A boundary-integral framework to simulate \\ the viscous erosion of a porous medium

### 2018-04-06 docs/methods/erosion_methods.tex
working on intro

The methods developed here will allow us to simulate the erosion of many bodies, as demonstrated in Fig.~\ref{fig:50bodies} for the case of 50 bodies (of randomly assigned initial sizes and positions). Here, the erosion of solid material arises from the mechanical stress induced by the surrounding flow. In particular, we take the local rate of erosion to be proportional to the magnitude of the shear stress---a law that has been verified by laboratory experiments~\cite{ris-moo-chi-she-zha2012}. As seen in Fig.~\ref{fig:50bodies}, erosion not only diminishes the size of the bodies, but also alters their shapes considerably. Notice, in particular, that many bodies develop corner-like features and nearly flat faces, with straight channels intervening between them. In the figure, color represents the vorticity of the surrounding flow, which reduces to shear on solid boundaries and thus provides a convenient tool for visualizing local erosion rates.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To solve the incompressible Stokes equations in the complex geometry, we
employ a boundary integral equation (BIE) method.
The present study develops numerical methods to simulate erosion of multiply-connected two-dimensional geometries. In this fashion, individual pores are resolved rather than homogenized.  A boundary integral equation (BIE) is used to solve the incompressible Stokes equations in the complex geometry.  The resulting method resolves individual grains and the non-negligible interactions between bodies.  To link the eroding body and fluid dynamics, the body is eroded in the normal direction at a rate that is proportional to the magnitude of the shear stress.

% Redundant statement: individual pores are resolved; individual grains are resolved.

To create high-fidelity and computationally efficient numerical methods,
two fast and accurate methods are employed.  First, the fluid velocity
and the erosion rate (shear stress) are computed with a
fast-multipole-accelerated Kryolv method applied to a spectrally
accurate discretization of the governing equations.  In particular,
GMRES~\cite{saa-sch1986} is applied to a spectrally-accurate Nystr\"om
discretization of the BIE~\cite{tre-wei2014}, a spectrally accurate
quadrature rule is used to compute the shear stress~\cite{sid-isr1988},
and all arclength derivatives are computed with Fourier differentiation.
In addition to computing the shear stress, we also compute the vorticity
in the fluid bulk.  The vorticity is particularly insightful since it is
identical to the shear stress on no-slip boundaries.  Therefore, by
computing the vorticity, we are able to identify regions where the
erosion rate will be fast or slow (Fig.~\ref{fig:50bodies}).

Once the erosion rate is computed, the shape is integrated in time with
second-order accuracy.  Tracking the interface of fluid-structure
interaction problems with a long time horizon is particularly
difficult---the mesh is often distorted and tangled.  Moreover, erosion
naturally leads to regions of high or infinite (i.e.~corners) curvature.
The interface tracking issues are resolved by, first, introducing a
tangential velocity field to maintain a grid that is equispaced in
arclength, but does not affect the dynamics, and, second, a tunable term
that penalizes regions of high curvature.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be handled in a numerically
stable fashion with an exponential integrator.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations} 
We begin by making two fundamental design choices: first, we restrict ourselves to two spatial dimensions, and second the bodies are considered immobile (i.e.~fixed to their initial locations). The restriction to two dimensions allows us to simulate the erosion of many bodies (up to $50$ in this manuscript) with high resolution, giving us the ability to accurately characterize the erosion of a porous medium. While BIEs are available in three dimensions, the extra computational cost limits the number of bodies that could be practically simulated, making porous-medium simulations less feasible. The second assumption of immobile bodies affords a simple framework to study the effects of erosion in isolation. While other effects, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically. We therefore defer the study of these effects to future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Related work} The effect of erosion on a geometry is quite
rich.  For instance, it has been hypothesized and numerically observed
that the erosion leads to a shear stress that is uniform on the body
which results in a uniform erosion rate.  The shape that has a uniform
shear stress is analyzed in three dimensions by Montenegro-Johnson and
Lauga~\cite{mon-lau2015} which is an extension of the analysis done by
Pironneau~\cite{pir1973} to find a Stokesian drag minimizing shape. 
%\todo[inline]{Double check this statement. I am not sure that the shape found by Lauga has uniform shear stress. The shape found by Pironneau does.}
In this idealized case, analytical axisymmetric shapes that minimize the
drag can be found.  However, the effect of erosion on a porous media is
less understood.

Simulating eroding bodies has been studied analytically, experimentally, and numerically.  Many of these works are motivated by high Reynolds number geophysical applications, such as the work of the second author~\cite{moo-ris-chi-zha-she2013, ris-moo-chi-she-zha2012, moore2017riemann} and others~\cite{par-izu2000, lag2000, coh-dev-sey-yi-szy-rot2015, han1969, Rothman2012, hewett2017evolution, ristroph2018sculpting, lopez2018cfd, lachaussee2018competitive, cohen2016erosion}.  The work most closely related to ours simulates erosion in a three-dimensional incompressible Stokesian fluid~\cite{mit-spa2016}, but they only consider single eroding body.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the paper} In Section~\ref{s:formulation}, we
describe the governing equations for both the fluid and the erosion
model in a \thL~framework.  In Section~\ref{s:method}, the numerical
methods in both space and time are described and Section~\ref{s:qoi}
describes how to compute the vorticity, pressure, and drag.
Section~\ref{s:SingleResults} gives several numerical examples for a
single body, and Section~\ref{s:MultiResults} presents numerical
examples for multiply-connected domains.  Finally, concluding remarks
are made in Section~\ref{s:conclusions}.

@@

Next, Figure \ref{fig:mbodyplots}(b) shows how the flow rate, $\umax$, changes in time as a result of fixing the pressure drop across the cell. Notice that, in this figure, the vertical axis is logarithmic. Both simulations show the flow rate to increase rapidly, even during the early stages of erosion when the area fraction has not changed that significantly. In fact, due to the logarithmic axis, the nearly linear trend seen in the figures corresponds to exponential growth of $\umax$. The change in area-fraction alone cannot account for this rapid growth, suggesting the reshaping process to be vital. In particular, the formations of channels observed in Fig.~\ref{fig:06bodies} is most likely the key feature that enables the higher flow rates.

>>>>

The methods developed here will allow us to simulate the erosion of many
bodies, as demonstrated in Fig.~\ref{fig:50bodies} for the case of 50
bodies (of randomly assigned initial sizes and positions). Here, the
erosion of solid material arises from the mechanical stress induced by
the surrounding flow. In particular, we take the local rate of erosion
to be proportional to the magnitude of the shear stress---a law that has
been verified by laboratory experiments~\cite{ris-moo-chi-she-zha2012}.
As seen in Fig.~\ref{fig:50bodies}, erosion not only diminishes the size
of the bodies, but also alters their shapes considerably. Notice, in
particular, that many bodies develop corner-like features and nearly
flat faces, with straight channels intervening between them. In the
figure, color represents the vorticity of the surrounding flow, which
reduces to shear on solid boundaries and thus provides a convenient tool
for visualizing local erosion rates. To perform this simulation of
Stokesian erosion of a porous media, two fast and accurate methods are
employed---one to solve the fluid equations and the other to integrate
the eroding grains.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The fluid equations are reformulated in terms of a boundary integral
equation (BIE).  Since the boundary conditions are Dirichlet, we use a
double-layer potential formulation so that the resulting BIE is
second-kind.  This second-kind BIE is discretized with the spectrally
accurate trapezoid rule~\cite{tre-wei2014} which results in a
well-conditioned dense linear system for an unknown density function.
This linear system is solved using fast-multipole-accelerated
GMRES~\cite{saa-sch1986}.  Once the density function is computed, the
erosion rate (shear stress) is computed with a spectrally accurate
odd-even quadrature rule~\cite{sid-isr1988}.  In addition, the pressure
and vorticity are computed in the fluid bulk (Fig.~\ref{fig:50bodies})
with an upsampled trapezoid rule in the fluid bulk to assist with
visualizing the effect of erosion in porous media.

%
%To create high-fidelity and computationally efficient numerical methods,
%two fast and accurate methods are employed.  First, the fluid velocity
%and the erosion rate (shear stress) are computed with a
%fast-multipole-accelerated Kryolv method applied to a spectrally
%accurate discretization of the governing equations.  In particular,
%GMRES~\cite{saa-sch1986} is applied to a spectrally-accurate Nystr\"om
%discretization of the BIE~\cite{tre-wei2014}, a spectrally accurate
%quadrature rule is used to compute the shear stress~\cite{sid-isr1988},
%and all arclength derivatives are computed with Fourier differentiation.
%In addition to computing the shear stress, we also compute the vorticity
%in the fluid bulk.  The vorticity is particularly insightful since it is
%identical to the shear stress on no-slip boundaries.  Therefore, by
%computing the vorticity, we are able to identify regions where the
%erosion rate will be fast or slow (Fig.~\ref{fig:50bodies}).

Once the erosion rate is computed, the grains are integrated in time with
second-order accuracy.  Tracking the interface of fluid-structure
interaction problems with a long time horizon is particularly
difficult---the mesh is often distorted and tangled.  Moreover, erosion
naturally leads to regions of high or infinite (i.e.~corners) curvature.
The interface tracking issues are resolved by, first, introducing a
tangential velocity field to maintain a grid that is equispaced in
arclength, but does not affect the dynamics, and, second, a tunable term
that penalizes regions of high curvature.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be handled in a numerically
stable fashion with an exponential integrator.

@@

The effect of erosion on a geometry is quite rich.  For instance, it has
been hypothesized and numerically observed that the erosion leads to a
shear stress that is uniform on the body which results in a uniform
erosion rate.  The shape that has a uniform shear stress is analyzed in
three dimensions by Montenegro-Johnson and Lauga~\cite{mon-lau2015}
which is an extension of the analysis done by Pironneau~\cite{pir1973}
to find a Stokesian drag minimizing shape. 
%\todo[inline]{Double check this statement. I am not sure that the shape found by Lauga has uniform shear stress. The shape found by Pironneau does.}
In this idealized case, analytical axisymmetric shapes that minimize the
drag can be found.  However, the effect of erosion on a porous media is
less understood.  The work most closely related to ours simulates erosion in a three-dimensional incompressible Stokesian fluid~\cite{mit-spa2016}, but they only consider single eroding body.

@@

We make two fundamental design choices: first, we restrict ourselves to two spatial dimensions, and second the bodies are considered immobile (i.e.~fixed to their initial locations). The restriction to two dimensions allows us to simulate the erosion of many bodies (up to $50$ in this manuscript) with high resolution, giving us the ability to accurately characterize the erosion of a porous medium. While BIEs are available in three dimensions, the extra computational cost limits the number of bodies that could be practically simulated, making porous-medium simulations less feasible. The second assumption of immobile bodies affords a simple framework to study the effects of erosion in isolation. While other effects, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically. We therefore defer the study of these effects to future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In Section~\ref{s:formulation}, we describe the governing equations for
both the fluid and the erosion model in a \thL~framework.  In
Section~\ref{s:method}, the numerical methods in both space and time are
described and Section~\ref{s:qoi} describes how to compute the
vorticity, pressure, and drag.  Section~\ref{s:SingleResults} gives
several numerical examples for a single body, and
Section~\ref{s:MultiResults} presents numerical examples for
multiply-connected domains.  Finally, concluding remarks are made in
Section~\ref{s:conclusions}.

@@

Next, Figure \ref{fig:mbodyplots}(b) shows how the flow rate, $\umax$, changes in time as a result of fixing the pressure drop across the cell. Notice that, in this figure, the vertical axis is logarithmic. Both simulations show the flow rate to increase rapidly, even during the early stages of erosion when the area fraction has not changed that significantly. Due to the logarithmic axis, the nearly linear trend seen in the figures corresponds to exponential growth of $\umax$. The change in area-fraction alone cannot account for this rapid growth, suggesting the reshaping process to be vital. In particular, the formations of channels observed in Fig.~\ref{fig:06bodies} is most likely the key feature that enables the higher flow rates.
